---
title: "COVID19Trends"
author: "reneemf"
date: "2020-09-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Which countries have had the highest number of positive cases against the number of tests?
*This project uses a cleaned Kaggle dataset [available here for download](https://dq-content.s3.amazonaws.com/505/covid19.csv)
  *Original dataset with regular updates [available here](https://www.kaggle.com/lin0li/covid19testing)
*The dataset used was collected between the 20th of January and the 1st of June 2020

```{r load packages and csv file}
library(readr)
covid_df <- read_csv("data/covid19.csv")
```
To begin we will analyze the dimensions and contents of the data

```{r covid_df analysis}

# View df dimensions
dim(covid_df)

# View df column names
vector_cols <- colnames(covid_df)
vector_cols

# View df head
head(covid_df)

# Glimpse df
library(tibble)
glimpse(covid_df)
```

### Dataset columns descriptions:
1. Date: Date
2. Continent_Name: Continent names
3. Two_Letter_Country_Code: Country codes
4. Country_Region: Country names
5. Province_State: States/province names; value is All States when state/provincial level data is not available
6. positive: *Cumulative* number of positive cases reported.
7. active: Number of actively cases on that **day**.
8. hospitalized: *Cumulative* number of hospitalized cases reported.
9. hospitalizedCurr: Number of actively hospitalized cases on that **day**.
10.recovered: *Cumulative* number of recovered cases reported.
11. death: *Cumulative* number of deaths reported.
12. total_tested: *Cumulative* number of tests conducted.
13. daily_tested: Number of tests conducted on the **day**; if daily data is unavailable, daily tested is averaged across number of days in between.
14. daily_positive: Number of positive cases reported on the **day**; if daily data is unavailable, daily positive is averaged across number of days in.

To focus on COVID cases and tests within entire countries, we will filter the data to retain only the rows where Province_State is listed as "All States"

```{r covid_df filtering}
library(dplyr)
covid_df_all_states <- covid_df %>% filter(Province_State == "All States") %>% select(-Province_State)
```

To study COVID cases and tests by country over time we will need to seperate columns describing cumulative data versus daily reports

* Selected cumulative data columns will be 
  * Date
  * Country_Region
  * positive
  * hospitalized
  * recovered
  * death
  * total_tested

* Selected daily reports data columns will be 
  * Date
  * Country_Region
  * active
  * hospitalizedCurr
  * daily_tested
  * daily_positive

```{r covid_df split into cumulative vs. daily}
covid_df_all_states_cumu <- covid_df %>% select(Date, Country_Region, positive, hospitalized, recovered, death, total_tested)
covid_df_all_states_daily <- covid_df %>% select(Date, Country_Region, active, hospitalizedCurr, daily_tested, daily_positive)
```

I plan to find overall number of COVID-19 tested, positive, active and hospitalized cases by country and then highlight the top ten countries

```{r summarize cases}
# Cumulative data columns
covid_df_all_states_cumu_sum <- covid_df_all_states_cumu %>% group_by(Country_Region) %>% summarize(positive = sum(positive), hospitalized = sum(hospitalized), recovered = sum(recovered), death = sum(death), tested = sum(total_tested), pos_v_test = positive/tested) 
covid_cumu_top_tests <- covid_df_all_states_cumu_sum %>% arrange(desc(tested))

# Daily reports data columns
covid_df_all_states_daily_sum <- covid_df_all_states_daily %>% group_by(Country_Region) %>% summarize(active = sum(active), hospitalized = sum(hospitalizedCurr), tested = sum(daily_tested), positive = sum(daily_positive), pos_v_test = positive/tested) 
covid_daily_top_tests <- covid_df_all_states_daily_sum %>% arrange(desc(tested))
```

```{r highlight top ten}
# Cumulative data columns
covid_cumu_top_10 <- head(covid_cumu_top_tests, 10)
covid_cumu_top_10

# Daily reports data columns
covid_daily_top_10 <- head(covid_daily_top_tests, 10)
covid_daily_top_10
```

Lets compare the numbers of positive cases against the number of tests

```{r positive cases vs. tests}
#Compare positive cases vs. tests for cumulative data columns
covid_cumu_top_pvt <- covid_df_all_states_cumu_sum %>% arrange(desc(pos_v_test))
covid_cumu_top_10_pvt <- head(covid_cumu_top_pvt, 10)
covid_cumu_top_10_pvt


#Compare positive cases vs. tests for daily reports data columns
covid_daily_top_pvt <- covid_df_all_states_cumu_sum %>% arrange(desc(pos_v_test))
covid_daily_top_10_pvt <- head(covid_daily_top_pvt, 10)
covid_daily_top_10_pvt
```
