---
title: "forestfires"
author: "reneemf"
date: "2020-09-15"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

## Introduction

```{r Load packages & Import data}
rm(list = ls()) # clear workspace

library(tidyverse)
library(ggthemes)

forest_fires <- read_csv('data/forestfires.csv')

summary(forest_fires)
head(forest_fires)
```

Data is taken from a 2006 forest fire prevention [publication](http://www3.dsi.uminho.pt/pcortez/fires.pdf) and can be found [here](https://archive.ics.uci.edu/ml/machine-learning-databases/forest-fires/)

When are forest fires most common? During which months? or days?

To begin analysis, mutate df to put months & days in chronological order 

```{r df mutation}
forest_fires <- forest_fires %>%
  mutate(month = factor(month, levels = c(tolower(month.abb))),
         day = factor(day, levels = c("mon", "tue", "wed", "thu", "fri", "sat", "sun")))
```

Set overall theme for all plots:
```{r plot theme}
plot_theme <- theme_hc() + theme(plot.title = element_text(family = "Helvetica", face = "bold", size = (15), hjust = 0.5),
                    axis.title = element_text(family = "Helvetica", size = (12)),
                    axis.text = element_text(family = "Helvetica", size = (12)), 
                    axis.ticks = element_line(colour = "grey70", size = 0.2),
                    panel.grid.major = element_line(colour = "grey70", size = 0.2),
                    panel.grid.minor = element_blank())
```

Plot forest fires per month
```{r fires per month}
fires_per_month <- forest_fires %>%
  group_by(month) %>%
  summarize(total_fires = n())

ggplot(fires_per_month, aes(x = month, y = total_fires)) +
  geom_bar(stat = 'identity') +
  ggtitle("Monthly Forest Fires in Portugal from \nJan 2000 to Dec 2003") +
  labs(y = 'Total Number of Fires', x = 'Month') +
  plot_theme + 
  ylim(0, 200)
```

Plot forest fires per day
```{r fires per day}
fires_per_day <- forest_fires %>%
  group_by(day) %>%
  summarize(total_fires = n())

ggplot(fires_per_day, aes(x = day, y = total_fires)) +
  geom_bar(stat = 'identity') +
  ggtitle("Daily Forest Fires in Portugal from \nJan 2000 to Dec 2003") +
  labs(y = 'Total Number of Fires', x = 'Day of the Week') +
  plot_theme +
  ylim(0, 100)
```

Fires seem to occur more in the late summer/early fall, especially August, and September followed by March. They allow occur more on the weekend (Sunday, Friday, and Saturday in order).


How are variables linked to forest fires, like temperature, rainfall, etc., distributed among the months and days measured?

Build function to make box plots
```{r box plot build}
# Build function to make box plots
boxplotter <- function(x, y) {
  ggplot(data = forest_fires) + 
    aes_string(x = x, y = y) +
    geom_boxplot() + plot_theme +
    ylim(0, 8)
}

# Assign variables
x_var_m <- names(forest_fires)[3] # month
x_var_d <- names(forest_fires)[4] # day
y_var <- names(forest_fires)[5:12] # "FFMC" "DMC"  "DC" "ISI" "temp" "RH" 
# "wind" and "rain" which stand for "Fine Fuel Moisture Code" 
# "Duff Moisture Code" "Drought Code", "Initial Spread Index", "Temperature", 
# "Relative Humidity (%)", "Wind", and "Rainfall"

# Visualize distributions by month and day
month_boxplot <- map2(x_var_m, y_var, boxplotter)
day_boxplot <- map2(x_var_d, y_var, boxplotter)
```

temp & DC (drought code, measures dry conditions) are both high in the summer & early fall months when more forest fires are occurring.

Are these late summer/early fall forest fires more or less severe than the fires throughout the rest of the year?


Forest fire size/spread will be used to assess severity

```{r scatter plot build}
# Build function to make scatter plots
scatterplotter <- function(x, y) {
  ggplot(data = forest_fires) + 
    aes_string(x = x, y = y) +
    geom_point() + plot_theme +
    ylim(0, 800)
}

# Assign variables
x_var_scat <- names(forest_fires)[5:12]
y_var_scat <- names(forest_fires)[13] #area

# Visualize which variables may affect forest fire size
scatterplots <- map2(x_var_scat, y_var_scat, scatterplotter)
scatterplots
```

Scatterplots do not show a very clear pattern, possibly due to outliers which will need to be addressed. Subsetting data may help in understanding how different variables contribute to fire spread (area) w/o the interference of outliers

```{r filtering}
# Outliers can be filtered out using interquartile range (IQR)
forest_fires_no_highs <- forest_fires %>%
  filter(area <= quantile(forest_fires$area, 0.75) + IQR(forest_fires$area) * 1.5)

# Rebuild function to make new scatter plots w/o v large values
scatterplotter_no_highs <- function(x, y) {
  ggplot(data = forest_fires_no_highs) + 
    aes_string(x = x, y = y) +
    geom_point() + plot_theme +
    ylim(0, 20)
}

sps_no_highs <- map2(x_var_scat, y_var_scat, scatterplotter_no_highs)
sps_no_highs
```

Plots have lots of data points clustered at the bottom. How would removing outliers from the bottom of the IQR affect them?

```{r rm outliers from IQR 0.25}
forest_fires_no_lows <- forest_fires %>%
  filter(area >= quantile(forest_fires$area, 0.25) - IQR(forest_fires$area) * 1.5)

# Rebuild function to make new scatter plots w/o v small values
scatterplotter_no_lows <- function(x, y) {
  ggplot(data = forest_fires_no_lows) + 
    aes_string(x = x, y = y) +
    geom_point() + plot_theme +
    ylim(0, 800)
}

sps_no_lows <- map2(x_var_scat, y_var_scat, scatterplotter_no_lows)
sps_no_lows
```

What is the impact of removing outliers on top & bottom of IQR?

```{r rm top & bottom IQR}
# Remove outliers on top & bottom of IQR
forest_fires_no_outs <- forest_fires %>%
  filter(area <= quantile(forest_fires$area, 0.75)+IQR(forest_fires$area)*1.5 & 
           area >= quantile(forest_fires$area, 0.25)-IQR(forest_fires$area)*1.5)

# Rebuild function to make new scatter plots w/o v small or v large values
scatterplotter_no_outs <- function(x, y) {
  ggplot(data = forest_fires_no_outs) + 
    aes_string(x = x, y = y) +
    geom_point() + plot_theme +
    ylim(0, 20)
}

sps_no_outs <- map2(x_var_scat, y_var_scat, scatterplotter_no_outs)
sps_no_outs
```

Plots look better but still have points clustered @ bottom 

forest_fires has entries of 0 as the area of some fires which seems suspect and is limiting how predictive my scatters can be. I will remove these

```{r rm zeros}
# Filter out 0s
forest_fires_no_zeros <- forest_fires %>%
  filter(area != 0)

# Rebuild function to make new scatter plots w/o 0s
scatterplotter_no_zeros <- function(x, y) {
  ggplot(data = forest_fires_no_zeros) + 
    aes_string(x = x, y = y) +
    geom_point() + plot_theme +
    ylim(0, 800)
}

sps_no_zeros <- map2(x_var_scat, y_var_scat, scatterplotter_no_zeros)
sps_no_zeros
```

How does removing both outliers and zeros affect the data?

```{r rm outliers & 0s}
# Plot data w/o outliers or 0s
forest_fires_no_outs_or_zeros <- forest_fires %>%
  filter(area != 0) %>%
  filter(area <= quantile(forest_fires$area, 0.75)+IQR(forest_fires$area)*1.5 & 
           area >= quantile(forest_fires$area, 0.25)-IQR(forest_fires$area)*1.5)

# Rebuild function to make new scatter plots w/o v small or v large values
scatterplotter_no_outs_or_zeros <- function(x, y) {
  ggplot(data = forest_fires_no_outs_or_zeros) + 
    aes_string(x = x, y = y) +
    geom_point() + plot_theme +
    ylim(0, 20)
}

sps_no_outs_or_zeros <- map2(x_var_scat, y_var_scat, 
                             scatterplotter_no_outs_or_zeros)
sps_no_outs_or_zeros
```

Honestly not seeing a clear pattern at this time 

Future plans: 
    compress all these plotting functions into one that takes any data set, any x & any y and put plots in the same screen(such as using the code: par(mfrow=c(n,n))), so they're easier to compare 
